---
title: "ASSIGNMENT 1"
author: "SAURABH MISHRA"
date: "2026-01-19"
output: html_document
---
##                         PREDICTIVE ANALYTICS
###                     Problem Set 1: An Introduction
Download “Boston” housing data from MASS library in R. Complete the task
given below and submit the report using R markdown. You need to copy each
question as well.
```{r}
library(MASS)
data(Boston)
```
## About the Boston data set
**Description of Variables in the Boston Housing Dataset**

The Boston housing dataset contains information on housing, demographic, and environmental characteristics of suburbs in Boston. The variables included in the dataset are described below:

crim - 	Per capita crime rate by town.

zn -	Proportion of residential land zoned for lots over 25,000 sq.ft.

indus- Proportion of non-retail business acres per town.

chas - 	Charles River dummy variable (1 if tract bounds river; 0 otherwise).

nox - 	Nitrogen oxides concentration (parts per 10 million).

rm - 	Average number of rooms per dwelling.

age - 	Proportion of owner-occupied units built prior to 1940.

dis	- Weighted mean of distances to five Boston employment centres.

rad	- Index of accessibility to radial highways.

tax	- Full-value property-tax rate per $10,000.

ptratio	- Pupil–teacher ratio by town.

black	- Transformed measure of the proportion of Black residents by town: 1000*(Bk-0.63)^2, where Bk is the actual proportion.

lstat	- Percentage of lower status population.

medv	- Median value of owner-occupied homes in $1000s.


**1. Report the “class” of the data set. How many rows and columns are in
this data set? What do the rows and columns represent?**
```{r}
class(Boston)
dim(Boston)
```
Each row represents a suburb of Boston.

Each column represents a variable describing housing, demographic, environmental, or socio-economic characteristics of the suburb.

**2. Create a smaller data set with the variables median value of owner-occupied
homes, per capita crime rate, nitrogen oxides concentration, proportion of
blacks and percentage of lower status of the population. Choosing median
value of owner occupied homes as the response and the rest as the pre-
dictors, make scatter plots of the response versus each predictor. Present
the scatter plots in different panels of the same graph. Comment on your
findings.**
```{r}
boston_small <- Boston[, c("medv", "crim", "nox", "black", "lstat")]
head(boston_small)
```
***Scatter Plot***
```{r}
par(mfrow = c(2,2))
plot(boston_small$crim, boston_small$medv, xlab="Crime Rate", ylab="Median Value", main="MEDV vs CRIM")
plot(boston_small$nox, boston_small$medv, xlab="NOX", ylab="Median Value", main="MEDV vs NOX")
plot(boston_small$black, boston_small$medv, xlab="Black", ylab="Median Value", main="MEDV vs BLACK")
plot(boston_small$lstat, boston_small$medv, xlab="LSTAT", ylab="Median Value", main="MEDV vs LSTAT")
```
***Findings***

1. Median house value decreases as crime rate increases.

2. A strong negative relationship exists between medv and lstat.

**3. Which suburb of Boston has lowest median value of owner-occupied homes?
What are the values of the other predictors mentioned in (2), for that
suburb. How do these values compare to the overall ranges for those pre-
dictors? Comment on your findings. Hint: Mention which percentile these
values belong to.**
```{r}
min_medv_index <- which.min(Boston$medv)
Boston[min_medv_index, c("medv", "crim", "nox", "black", "lstat")]
```
```{r}
# Percentiles of predictors for that suburb
suburb_values <- Boston[min_medv_index, c("crim", "nox", "black", "lstat")]
percentiles <- sapply(
  c("crim", "nox", "black", "lstat"),
  function(var) ecdf(Boston[[var]])(suburb_values[[var]])
)

percentiles


```
***Comments***

The suburb corresponding to 399th row has lowest median value of owner-occupied homes.

The suburb with the lowest median house value lies in the upper percentiles of crime rate and lower-status population.


Pollution levels are also relatively high.

These characteristics explain the very low housing prices in this suburb.

**4. Does any suburb of Boston stand out for having notably high crime rates,
tax rates, or pupil–teacher ratios? Hint: Use a boxplot to detect any
outliers. If so, identify the suburbs that show the outlier values.**
```{r}
par(mfrow = c(1, 3))

boxplot(Boston$crim, main = "Crime Rate", ylab = "CRIM")
boxplot(Boston$tax, main = "Tax Rate", ylab = "TAX")
boxplot(Boston$ptratio, main = "Pupil–Teacher Ratio", ylab = "PTRATIO")

```

```{r}
# Identify suburbs with outliers
out_crim <- which(Boston$crim %in% boxplot.stats(Boston$crim)$out)
out_tax <- which(Boston$tax %in% boxplot.stats(Boston$tax)$out)
out_ptratio <- which(Boston$ptratio %in% boxplot.stats(Boston$ptratio)$out)

out_crim
out_tax
out_ptratio

```
***Comments***

A few suburbs exhibit extremely high crime rates.

Some suburbs also show unusually high tax rates and pupil–teacher ratios.

These outliers may strongly influence predictive models and should be handled carefully.
